---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(here)
```


Using merged loops files from MGG and LD.
H3K27Ac HiChIP in primary cells

```{r}
files = list.files(path = here("data/mergedloops/loops/"),
           full.names = T)
cell_types = str_remove(files, here("data/mergedloops/loops//"))
cell_types = str_remove(cell_types, "\\..*$")
```


Make the set of promoters
```{r}
gene_table <- read_tsv(here("data/ucsc_resfseq_refseqall"),
                       col_types = "ccccnncccccccccc")
# this file comes from UCSC table browser
# group: genes and gene predictions
# track: NCBI RefSeq
# table: RefSeq All (ncbiRefSeq)
# hg37
# output: all fields from selected table

# gene_conversion <- read_tsv("~/Documents/MPRA/Cancer_MPRA_Scaleup/eGeneFiles/HiChIP_112820/ucsc_id_conversion.tsv",
 #                           col_types = "cccccccccc")

# gene_table_merge <- left_join(gene_table, gene_conversion, by=c("#name"="#kgID"))
gene_table_merge <- gene_table
gene_table_merge <- dplyr::filter(gene_table_merge, !grepl("_", chrom))
gene_table_flip <- gene_table_merge %>% mutate(promoter = ifelse(strand=="+", txStart, txEnd))
  
gene_table_flip <- dplyr::select(gene_table_flip, name2, chrom, promoter) %>% distinct()

promoters <- gene_table_flip %>% group_by(name2, chrom) %>%
  summarise(prom_start = min(promoter) - 5000,
            prom_end = max(promoter) + 5000)
promoters <- promoters %>% mutate(chrom = str_remove(chrom, "chr"))
colnames(promoters)[1] <- "geneSymbol"
```

Make the set of SNPs
```{r}
res_merge <- read_tsv(here("output/res_merge_gtex.tsv"))
snp_positions <- dplyr::select(res_merge, Causal_SNP, Chr, Start) %>% distinct()
snp_positions$Start <- as.numeric(snp_positions$Start)
```



Read in the hichip file
Make a distinct anchor list
Annotate anchor list with snps
Filter loop list down to those that touch snps
Annotate anchor list with promoters
Combine the annotations
Filter to loops that have snps and genes
Add cell type

Start the for loop:

```{r}
loops_all <- data.frame()
for (cell in cell_types){
  print(cell)
  loops_file <- read_csv(files[grep(cell, files)])
  loops_file <- loops_file %>% dplyr::select(-`...1`, -count) %>%
    separate(source, into=c("chr1", "s1", "e1"), remove=F) %>%
    separate(target, into=c("chr2", "s2", "e2"), remove=F)
  loops_file <- loops_file %>% mutate_at(c("s1", "e1", "s2", "e2"), as.numeric)
  loops_file <- loops_file %>% mutate(chr1 = str_remove(chr1, "chr"),
                                      chr2 = str_remove(chr2, "chr"))
  
  # Make a distinct anchor list
  anchors_1 <- distinct(dplyr::select(loops_file, source, chr1, s1, e1))
  colnames(anchors_1) <- c("anchor", "chrom", "start", "end")
  anchors_2 <- distinct(dplyr::select(loops_file, target, chr2, s2, e2))
  colnames(anchors_2) <- c("anchor", "chrom", "start", "end")
  anchors <- distinct(rbind(anchors_1, anchors_2))
  
  # Annotate anchor list with snps
  check_snp <- function(snp, snp_chrom, snp_pos, anchors) {
    anchor_snps <- dplyr::filter(anchors, (chrom == as.character(snp_chrom)) &
                                   (as.numeric(start) < as.numeric(snp_pos) & as.numeric(snp_pos) < as.numeric(end)))
    anchor_snps <- anchor_snps
    anchor_snps$snp <- unname(snp)
    anchor_snps$snp_pos <- unname(snp_pos)
    return(as.data.frame(anchor_snps))}
  
  
  check_snp <- function(snp, snp_chrom, snp_pos, anchors) {
    anchor_snps <- dplyr::filter(anchors, (chrom == as.character(snp_chrom)) &
                                   (as.numeric(start) < as.numeric(snp_pos) & as.numeric(snp_pos) < as.numeric(end)))
    anchor_snps <- anchor_snps %>% mutate(snp = snp, snp_pos = snp_pos)
    return(as.data.frame(anchor_snps))}
  
  # test <- check_snp("rs59351655", "8", 128351048, anchors)
  
  # check if apply is the problem
  snps_checked = data.frame()
  for (i in 1:dim(snp_positions)[1]){
    x = snp_positions[i,]
    snp_checked = check_snp(x["Causal_SNP"], x["Chr"], x["Start"], anchors)
    if (dim(snp_checked)[1] > 0){
      if(dim(snps_checked)[1] > 0){
        snps_checked = base::rbind(as.matrix(snps_checked), as.matrix(snp_checked))
      }else{
        snps_checked = snp_checked
      }
    }
  }
  anchor_snp_filt_df <- as.data.frame(snps_checked)
  
  anchor_snp_filt_df <- anchor_snp_filt_df %>% group_by(anchor, chrom, start, end) %>%
    summarise(snp = paste(unique(snp), collapse=";"))
  
  # Filter loop list down to those that touch snps
  loops_filt <- dplyr::filter(loops_file, source %in% anchor_snp_filt_df$anchor |
                                target %in% anchor_snp_filt_df$anchor)
  anchors_filt <- anchors[anchors$anchor %in% loops_filt$source |
                            anchors$anchor %in% loops_filt$target,]
  
  # change gene check to for loop too
  
  # Annotate anchor list with promoters
  check_prom <- function(gene, promoter_chrom, promoter_start, promoter_end, anchors) {
    anchor_genes <- dplyr::filter(anchors, (as.character(chrom) == as.character(promoter_chrom)) &
                                    ((as.numeric(start) < as.numeric(promoter_start) & 
                                        as.numeric(promoter_start) < as.numeric(end)) |
                                       (as.numeric(start) < as.numeric(promoter_end) & 
                                          as.numeric(promoter_end) < as.numeric(end)) |
                                       (as.numeric(promoter_start) < as.numeric(start) &
                                          as.numeric(end) < as.numeric(promoter_end))))
    if(dim(anchor_genes)[1] > 0){
      anchor_genes$gene <- gene
      anchor_genes <- anchor_genes %>% 
        mutate_at(c("anchor", "chrom"), as.character)
    }
    return(as.data.frame(anchor_genes))}
  
  genes_checked = data.frame()
  for (i in 1:dim(promoters)[1]){
    x = promoters[i,]
    gene_checked = check_prom(x[["geneSymbol"]], x[["chrom"]], x[["prom_start"]], x[["prom_end"]],
                              anchors_filt)
    if (dim(gene_checked)[1] > 0){
      if(dim(genes_checked)[1] > 0){
        genes_checked = base::rbind(as.matrix(genes_checked), as.matrix(gene_checked))
      }else{
        genes_checked = gene_checked
      }
    }
  }
  
  
  anchor_prom_filt_df <- as.data.frame(genes_checked)
  anchor_prom_filt_df <- anchor_prom_filt_df %>% group_by(anchor, chrom, start, end) %>%
    summarise(gene = paste(unique(gene), collapse=";"))
  
  # Combine the annotations
  anchors_annot <- full_join(anchor_snp_filt_df, anchor_prom_filt_df)
  
  # Filter to loops that have snps and genes
  loops_filt <- dplyr::filter(loops_filt, source %in% anchors_annot$anchor &
                                target %in% anchors_annot$anchor)
  loops_filt <- distinct(dplyr::select(loops_filt, source, target))
  loops_annot <- left_join(loops_filt, dplyr::select(ungroup(anchors_annot), anchor, snp, gene),
                           by = c("source"="anchor"))
  loops_annot <- left_join(loops_annot, dplyr::select(ungroup(anchors_annot), anchor, snp, gene),
                           by = c("target"="anchor"))
  loops_annot <- dplyr::filter(loops_annot, !(is.na(gene.x) & is.na(gene.y)))
  loops_annot$cell_type <- cell
  if(dim(loops_all)[1] > 0){
    loops_all = rbind(loops_all, loops_annot)
  } else{
    loops_all = loops_annot
  }
}
```

Combine GDS samples - these are from different stages of keratinocyte differentiation (Day 0, Day 3 and Day 6)
```{r}
# removing GMs since no longer in the cancer MPRA analysis
loops_all <- dplyr::filter(loops_all, !grepl("GM", cell_type))
loops_all <- loops_all %>% 
  mutate(cell_type = case_when(
    grepl("GDS", cell_type) ~ "kc",
    grepl("Air", cell_type) ~ "airway",
    grepl("Ast", cell_type) ~ "ast",
    grepl("Colon", cell_type) ~ "colon",
    grepl("Uter", cell_type) ~ "endo",
    grepl("Eso", cell_type) ~ "eso",
    grepl("GM", cell_type) ~ "gm",
    grepl("HMEC", cell_type) ~ "hmec",
    grepl("Mel", cell_type) ~ "mc",
    grepl("Ov", cell_type) ~ "ov",
    grepl("Panc", cell_type) ~ "panc",
    grepl("Pros", cell_type) ~ "pros",
    grepl("Renal", cell_type) ~ "renal",
    grepl("Thy", cell_type) ~ "thy"))

```


Separate out into loops where SNP and promoter are in different ends and those where it's the same
```{r}
loops_unnest <- loops_all %>% mutate(snp.x = str_split(snp.x, ";")) %>%
    unnest(cols=c(snp.x))
loops_unnest <- loops_unnest %>% mutate(snp.y = str_split(snp.y, ";")) %>%
    unnest(cols=c(snp.y))
loops_x = dplyr::select(loops_unnest, snp.x, gene.y, cell_type) %>%
  dplyr::filter(!is.na(snp.x)) %>% distinct()
loops_y = dplyr::select(loops_unnest, snp.y, gene.x, cell_type) %>%
  dplyr::filter(!is.na(snp.y)) %>% distinct()
colnames(loops_x) <- c("snp", "gene", "cell_type")
colnames(loops_y) <- c("snp", "gene", "cell_type")
loops <- rbind(loops_x, loops_y)
loops <- distinct(loops)
loops_snpcell <- loops %>% group_by(snp, cell_type) %>%
  summarise(gene = paste(unique(gene)[!is.na(unique(gene))], collapse=";"))
loops_snpcell <- dplyr::filter(loops_snpcell, !gene == "")
```

```{r}
write_tsv(loops_snpcell, here("output/hichiploops_looponly_snpcellgrouped_20230626.tsv"))

```


```{r}
same_bin_x = dplyr::select(loops_unnest, snp.x, gene.x, cell_type) %>%
  dplyr::filter(!is.na(snp.x) & !is.na(gene.x)) %>% distinct()
same_bin_y = dplyr::select(loops_unnest, snp.y, gene.y, cell_type) %>%
  dplyr::filter(!is.na(snp.y) & !is.na(gene.y)) %>% distinct()
colnames(same_bin_x) <- c("snp", "gene", "cell_type")
colnames(same_bin_y) <- c("snp", "gene", "cell_type")
same_bin <- rbind(same_bin_x, same_bin_y)
same_bin <- distinct(same_bin)
same_bin_snpcell <- same_bin %>% group_by(snp, cell_type) %>%
  summarise(gene = paste(unique(gene)[!is.na(unique(gene))], collapse=";"))
same_bin_snpcell <- dplyr::filter(same_bin_snpcell, !gene == "")

```

```{r}
write_tsv(same_bin_snpcell, here("output/hichip_samebin_snpcellgrouped_20230626.tsv"))

```

Go through and mark connections in the relevant cell type
```{r}
snp_disease <- res_merge %>% dplyr::filter(!is.na(disease)) %>%
  dplyr::select(Causal_SNP, disease)

```

```{r}
loops_snpcell$loop <- T
same_bin_snpcell$loop <- F
hichip <- loops_snpcell
```

```{r}
hichip <- left_join(hichip, dplyr::select(snp_disease, Causal_SNP, disease), by=c("snp"="Causal_SNP"))
hichip <- dplyr::filter(hichip, !(disease==""))
hichip <- hichip %>% rowwise() %>% mutate(same_cell_type = grepl(cell_type, disease))
```

```{r}
hichip_correct_celltype <- hichip %>% dplyr::filter(same_cell_type == T)
hichip_correct_celltype <- hichip_correct_celltype %>% group_by(snp) %>%
  summarise(hichip_correct_cell = paste(unique(gene), collapse=","),
            hichip_correct_cell_types = paste(unique(cell_type), collapse=","))
hichip_all_cell <- hichip %>% group_by(snp) %>%
  summarise(hichip_all = paste(gene, collapse=","),
            hichip_all_cell = paste(cell_type, collapse=","))
```


```{r}
res_merge_hichip <- left_join(res_merge, hichip_correct_celltype,
                              by=c("Causal_SNP"="snp"))
res_merge_hichip <- dplyr::select(res_merge_hichip, Causal_SNP,  
                                  disease, hichip_egene = hichip_correct_cell, 
                                  hichip_cell_type = hichip_correct_cell_types)

```

Find genes within 5kb - old version.
```{r}
gene_loc <- dplyr::select(gene_table_merge, chrom, txStart, txEnd, name2)
gene_loc_gp <- gene_loc %>% group_by(name2) %>%
  summarise(chrom=unique(chrom),
            txStart = min(txStart),
            txEnd = max(txEnd))
# gene_loc <- gene_loc_gp %>% mutate(start = txStart - 10000,
#                                end = txEnd + 10000) 
gene_loc <- gene_loc_gp %>% mutate(start = txStart - 5000,
                                end = txEnd + 5000) 
gene_loc <- gene_loc %>% mutate(start = ifelse(start < 0, 0, start)) 
gene_loc <- gene_loc %>% mutate(chrom = str_remove(chrom, "chr"))
find_nearby <- function(snp, snp_chrom, snp_pos, gene_loc) {
  gene_snps <- dplyr::filter(gene_loc, (chrom == snp_chrom) &
                                (as.numeric(start) < as.numeric(snp_pos) & as.numeric(snp_pos) < as.numeric(end)))
  gene_snps$snp <- unname(snp)
  gene_snps$snp_pos <- unname(snp_pos)
  return(as.data.frame(gene_snps))}


nearby_genes_5k = data.frame()
for (i in 1:dim(snp_positions)[1]){
  x = snp_positions[i,]
  nearest_genes_snp = find_nearby(x[["Causal_SNP"]], x[["Chr"]], x[["Start"]], gene_loc)
  if (dim(nearest_genes_snp)[1] > 0){
      if(dim(nearby_genes_5k)[1] > 0){
    nearby_genes_5k = base::rbind(as.matrix(nearby_genes_5k), as.matrix(nearest_genes_snp))
  }else{
    nearby_genes_5k = nearest_genes_snp
  }
  }
}
nearby_genes <- as.data.frame(nearby_genes_5k)

# using nearest_genes with 10kb filter
nearby_grouped <- nearby_genes %>% group_by(snp) %>%
  summarise(nearby_genes = paste(sort(unique(name2)), collapse=","))
  
  

```


Find nearest gene

```{r}
find_nearest <- function(snp_df, gene_loc) {
  gene_snps <- dplyr::filter(gene_loc, (chrom == snp_df$chrom))
  gene_snps <- rbind(gene_snps, snp_df)
  gene_snps <- gene_snps[order(as.numeric(gene_snps$pos)),]
  snp_ind <-  grep(snp_df$id, gene_snps$id)
  two_gene <- gene_snps[c((snp_ind-1),(snp_ind+1)),]
  two_gene <- two_gene %>% 
    mutate(snp_dist = abs(pos - snp_df$pos))
  winner <- data.frame(snp = snp_df$id, 
                       nearest_gene = two_gene$id[which.min(two_gene$snp_dist)])
  return(winner)}

gene_nearest_reformat <- dplyr::filter(gene_table, !grepl("_", chrom))
# flip so it's closest promoter
gene_nearest_reformat <- gene_nearest_reformat %>% mutate(pos = ifelse(strand=="+", txStart, txEnd))
gene_nearest_reformat <- dplyr::select(gene_nearest_reformat, chrom, pos, id = name2)
gene_nearest_reformat <- gene_nearest_reformat %>%
  mutate(chrom = str_remove(chrom, "chr"))


snp_nearest_reformat <- snp_positions %>%
  rename(id = Causal_SNP, chrom = Chr, pos = Start)

nearest_gene = data.frame()

for (i in 1:dim(snp_positions)[1]){
  x = snp_nearest_reformat[i,]
  nearest_genes_snp = find_nearest(x, gene_nearest_reformat)
  if (dim(nearest_genes_snp)[1] > 0){
      if(dim(nearest_gene)[1] > 0){
    nearest_gene = base::rbind(as.matrix(nearest_gene), as.matrix(nearest_genes_snp))
  }else{
    nearest_gene = nearest_genes_snp
  }
  }
}
nearest_gene_df <- as.data.frame(nearest_gene)
nearest_gene_df <- distinct(nearest_gene_df)
nearest_gene_df <- nearest_gene_df %>% rename(Causal_SNP = snp)


```

Add nearby genes, nearest gene and hichip to res_merge (table of MPRA results and background)

```{r}
colnames(hichip_correct_celltype) <- c("Causal_SNP", "hichip_egene", "hichip_cell_type")
res_merge_hichip <- left_join(res_merge,
                              hichip_correct_celltype)
colnames(nearby_grouped) <- c("Causal_SNP", "genes_within_10kb")
colnames(nearest_gene_df) <- c("Causal_SNP", "nearest_gene")
res_merge_gene <- left_join(res_merge_hichip, nearby_grouped)
res_merge_gene <- left_join(res_merge_gene, nearest_gene_df)

check <-   dplyr::select(res_merge_gene, Causal_SNP, hichip_egene, genes_within_10kb,
                         nearest_gene)                                       
                                            
```


```{r}
write_tsv(res_merge_gene, here("output/res_merge_gtex_hichip_proximity.tsv"))
```


