---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
```

Identify which SNVs are disease specific hits (ie, differentially active in the MPRA in the cell type relevant to the cancer they're associated with)
Recode column disease_expanded so it's using the same words for each cancer/tissue type.
Add back GWAS lib info
```{r}
res_merge <- read_tsv(here("output/res_merge_gtex_hichip_proximity_eqtlgen.tsv"))
```

```{r}
# renaming some columns to match other analyses
res_merge <- res_merge %>%
  rename(Chr_37 = Chr,
         Start_37 = Start,
         gtex_egenes_all = gtex_egene_all,
         gtex_egenes_spec = gtex_egene_spec,
         eqtlgen_egenes = eqtlgen_egene)


```
Leaving out: lead_annotation_problem, haploreg_motifs, disease_subthresh, atac_tissues
Need: leadSNP_PUBMEDID, leadSNP_OR.or.BETA, STUDY, 
Maybe: study_type
Check: gtex_allele_match (is this covered in alt?)


Annotate hits for any cell type (SNVs differentially active in that cell type) and hits in the correct cell type
```{r}
hits <- res_merge[, grepl("fdr", colnames(res_merge))]
hits$locus <- res_merge$locus
hits$disease <- res_merge$disease
hits <- pivot_longer(hits, cols= airway_fdr:thy_fdr, names_to = "cell_type", values_to="fdr")
hits <- hits %>% mutate(cell_type = str_remove(cell_type, "_fdr") )
hits$fdr <- as.numeric(hits$fdr)

hits <- hits %>% mutate(hit_any_celltype = ifelse(fdr < 0.05, cell_type,NA))
hits <- hits %>% rowwise() %>%
  mutate(hit_dis_spec = ifelse(grepl(hit_any_celltype, disease), cell_type, NA))
hits <- hits %>% group_by(locus) %>%
  summarise(hit_any_celltype = paste(sort(unique(hit_any_celltype[!is.na(hit_any_celltype)])),
                                     collapse=","),
            hit_spec = paste(sort(unique(hit_dis_spec[!is.na(hit_dis_spec)])), collapse=","))
```

```{r}
res_merge_hits <- left_join(res_merge, hits)
```


```{r}
 write_tsv(res_merge_hits, here("output/res_merge_withhitannot_20230627.tsv"))

```

```{r}
old_res_merge <- read_tsv("~/Documents/Old_CancerMPRA_Rproj/CancerMPRA_noGM_noH9/output/res_merge_withhitannot_20230627.tsv")

test <- lapply(1:72, function(x) sum((is.na(old_res_merge[[x]]) & is.na(res_merge_hits[[x]])) |
                                       old_res_merge[[x]] == res_merge_hits[[x]], na.rm=T))
names(test) <- colnames(old_res_merge)

check <- full_join(dplyr::select(old_res_merge, locus, disease, gtex_egenes_all, 
                                 gtex_egenes_spec, gtex_spec_lead_shared),
                   dplyr::select(res_merge_hits, locus, gtex_egenes_all, 
                                 gtex_egenes_spec, gtex_spec_lead_shared),
                   by = c("locus"))
check2 <- dplyr::filter(check, !gtex_spec_lead_shared.x == gtex_spec_lead_shared.y)

check <- full_join(dplyr::select(old_res_merge, locus, disease, airway_fdr),
                   dplyr::select(res_merge_hits, locus, airway_fdr),
                   by = c("locus"))
check2 <- dplyr::filter(check, !airway_fdr.x == airway_fdr.y)

sum(!res_merge_hits$hit_spec == "")
length(unique)

```



