---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(here)
library(tidyverse)
library(MPRAnalyze)
library(BiocParallel)
```


MPRAnalyze results were analyzed using MPRAnalyze 1.8.0. It's inelegant, but to run this I use rswitch to go back to R version 4.0.2, where I have BiocManager 1.30.20, BiocVersion 3.12.0 and MPRAnalyze 1.8.0 installed. 

Analyze the MPRA counts files using MPRAnalyze quantitative mode
Gives an estimate of transcription rate for each oligo

Read in counts file
```{r}
lib <- read_tsv(here("data/nextseq_counts_111920.tsv")) %>%
  dplyr::select(-gm_1, -gm_2, -gm_3, -h9_1, -h9_2, -h9_3)

```

Separate oligo name into locus, allele, barcode
```{r}
lib <- lib %>% separate(oligo, into=c("oligo_type", "locus", "allele", "barcode"), 
                        sep="_", remove=F) %>%
  mutate(locus = paste(oligo_type, locus, sep="_")) %>%
  dplyr::select(-oligo, -oligo_type)
```

For quantitative analysis, keep scrambles and motifs
Don't split on alleles

Make DNA matrix for MPRAnalyze
```{r}
lib <- lib %>% unite(locus, locus, allele)

dna <- lib %>% dplyr::select(locus, barcode, plasmid_2, plasmid_3)%>%
  pivot_longer(cols=c(plasmid_2, plasmid_3), names_to = "dna_batch", 
               values_to = "count") %>%
  mutate(dna_batch = ifelse(dna_batch == "plasmid_2", "rep1", "rep2")) %>% 
  rowwise() %>%
  mutate(batch.bc = paste(dna_batch, barcode, sep="-")) %>%
  dplyr::select(-barcode, -dna_batch) %>%
  spread(key= batch.bc, value = count)
future_rownames <- dna$locus
dna <- dna %>% ungroup() %>% dplyr::select(-locus)
dna[is.na(dna)] <- 0
dna <- as.matrix(dna)
rownames(dna) <- future_rownames

```

Make RNA matrix with all celltypes
```{r}
rna <- lib %>% dplyr::select(-plasmid_2, -plasmid_3) %>%
  pivot_longer(c(-locus, -barcode), names_to = "cell_type",
               values_to="count")
rna <- rna %>% unite("cell_type", cell_type, barcode, sep="-")
rna <- rna %>% pivot_wider(names_from=cell_type, values_from = count)

future_rownames <- rna$locus
rna <- rna %>% ungroup() %>% dplyr::select(-locus)
rna[is.na(rna)] <- 0
rna <- as.matrix(rna)
rownames(rna) <- future_rownames
rna <- rna[order(as.character(rownames(rna))),]

```

Make DNA column annotations
```{r}
col_dna = data.frame(dna_batch = str_remove(colnames(dna), "-.*$"),
                      barcode = str_remove(colnames(dna), "^.*-"))
rownames(col_dna) <- colnames(dna)
```

```{r}
cells = unique(str_remove(colnames(rna), "_.*"))
res = data.frame()
loci = distinct(dplyr::select(lib, locus))

# dna and col_dna can stay the same
# rna needs to be filtered for only the cell related columns
# rna col annot needs to be made
rna_cell <- rna

control_snps <- grep("scr", rownames(rna_cell))
col_rna = data.frame(condition = str_remove(colnames(rna_cell), "_.*"),
                     batch = str_remove(str_remove(colnames(rna_cell), "^.*_"),
                                        "-.*$"),
                     barcode = str_remove(str_remove(colnames(rna_cell), "-ref$|-alt$"), ".*-"))
rownames(col_rna) <- colnames(rna_cell)
obj <- MpraObject(dnaCounts =dna, rnaCounts =rna_cell, dnaAnnot = col_dna, 
                  rnaAnnot = col_rna, controls = control_snps)
obj <- estimateDepthFactors(obj, lib.factor = "dna_batch", which.lib = "dna", 
                            depth.estimator = "uq")
obj <- estimateDepthFactors(obj,which.lib = "rna", lib.factor= c("batch","condition"),
                            depth.estimator = "uq")
obj <- analyzeQuantification(obj = obj, 
                             dnaDesign = ~ barcode, 
                             rnaDesign = ~ condition)
  
alpha <- getAlpha(obj, by.factor = "all")

colnames(alpha)[1] <- "conditionairway"
mpraq_full = data.frame()
for (conditioncell in colnames(alpha)){
  cell <- str_remove(conditioncell, "^condition")
  res_df <- testEmpirical(obj = obj, statistic = alpha[[conditioncell]])
  res_df$locus <- rownames(alpha)
  res_df$celltype <- cell
  write_tsv(res_df, paste0(here("output/MPRAnalyze_quantitative_intermediates/"),
                        cell, "_mpranalyze_quantitative.tsv"))
  mpraq_full = bind_rows(mpraq_full, res_df)
}

```

```{r}
mpraq_full <- mpraq_full %>% mutate(oligo_type = str_remove(locus, "^.*_"),
                                    oligo_id = locus,
                                    locus = str_remove(oligo_id, "_ref|_alt"))

lib_bg <- read_tsv(here("output/cancer_lib_haploreg_diseasereannotated.tsv"))
res_cut <- dplyr::select(lib_bg, locus, Causal_SNP, Ref, Alt, snp_sequence) %>% distinct()

mpraq_ext <- left_join(mpraq_full, res_cut)
colnames(mpraq_ext)
mpraq_ext <- dplyr::select(mpraq_ext, celltype, locus, oligo_id, oligo_type, Causal_SNP,  statistic, control, zscore, mad.score, pval.mad,
                            pval.zscore, pval.empirical, Ref, Alt, snp_sequence)

mpraq_ext <- mpraq_ext %>% 
  mutate(fdr = p.adjust(pval.mad, method = "fdr"))
write_tsv(mpraq_ext, here("output/mpranalyze_quantitative_results.tsv"))

```


