---
title: "MPRAnalyze quantitative activity plots"
output: html_document
date: "2023-06-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(here)
```

Activity plots - plot barplots of transcriptional activity for different groups of SNVs, and a summary of how many were active in each cell type.

Want to show: number significantly active, separately for each cell types
Possible groups:
Motif
Positive
Scramble
SNPs associated with that cell type
All SNPs


```{r}
# mpranalyze quantitative results
mpraq <- read_tsv(here("output/mpranalyze_quantitative_results.tsv"))
res_merge <- read_tsv(here("output/res_merge_withhitannot_20230627.tsv"))
name_key <- read_tsv(here("data/celltype_name_colors.txt"))
```

```{r}
mpraq <- left_join(mpraq, distinct(dplyr::select(res_merge, locus, disease)))
```


```{r}
mpraq_plot <- mpraq %>% 
  mutate(seq_type = case_when(oligo_type == "motif" ~ "motif",
                   oligo_type == "scramble" ~ "scramble",
                   oligo_type %in% c("alt", "ref") ~ "SNP"))
mpraq_plot <- mpraq_plot %>%
  mutate(active_seq = ifelse(fdr < 0.05, "Active", "Inactive"))
mpraq_plot$active_seq <- factor(mpraq_plot$active_seq, 
                                levels = c("Inactive", "Active"),
                     ordered = T)

mpraq_plot <- mpraq_plot %>% 
  mutate(seq_type_2 = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    oligo_type == "motif" ~ "motif",
                   oligo_type == "scramble" ~ "scramble",
                   oligo_type %in% c("alt", "ref") ~ "SNP"))
```



```{r}
cell_sep <- mpraq_plot %>% 
  dplyr::select(-control, -zscore, -mad.score, -pval.mad,
                -pval.zscore, -pval.empirical, -Ref, -Alt, -snp_sequence,
                -fdr, -seq_type, -seq_type_2)

cell_sep <- cell_sep %>%
  mutate(hmec_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("hmec", disease) ~ "HMEC",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    airway_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("airway", disease) ~ "Airway",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    thy_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("thy", disease) ~ "Thyroid",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    colon_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("colon", disease) ~ "Colon",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    pros_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("pros", disease) ~ "Prostate",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    mc_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("mc", disease) ~ "MC",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    endo_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("endo", disease) ~ "Endometrial",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    ov_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("ov", disease) ~ "Ovarian",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    eso_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("eso", disease) ~ "Esophageal",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    kc_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("kc", disease) ~ "KC",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    panc_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("panc", disease) ~ "Pancreatic",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    renal_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("renal", disease) ~ "Renal",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    ast_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    grepl("ast", disease) ~ "Astrocyte",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"),
    
    lx_groups = case_when(
    grepl("control_JM", locus) ~ "pos_control",
    oligo_type == "motif" ~ "motif",
    oligo_type == "scramble" ~ "scramble",
    oligo_type %in% c("alt", "ref") ~ "SNP"))


cell_sep$hmec_groups <- factor(cell_sep$hmec_groups, 
                           levels = c("HMEC", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$airway_groups <- factor(cell_sep$airway_groups, 
                           levels = c("Airway", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$thy_groups <- factor(cell_sep$thy_groups, 
                           levels = c("Thyroid", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$colon_groups <- factor(cell_sep$colon_groups, 
                           levels = c("Colon", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$pros_groups <- factor(cell_sep$pros_groups, 
                           levels = c("Prostate", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$mc_groups <- factor(cell_sep$mc_groups, 
                           levels = c("MC", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$endo_groups <- factor(cell_sep$endo_groups, 
                           levels = c("Endometrial", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$ov_groups <- factor(cell_sep$ov_groups, 
                           levels = c("Ovarian", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$eso_groups <- factor(cell_sep$eso_groups, 
                           levels = c("Esophageal", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$kc_groups <- factor(cell_sep$kc_groups, 
                           levels = c("KC", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$panc_groups <- factor(cell_sep$panc_groups, 
                           levels = c("Pancreatic", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$renal_groups <- factor(cell_sep$renal_groups, 
                           levels = c("Renal", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$ast_groups <- factor(cell_sep$ast_groups, 
                           levels = c("Astrocyte", "SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)
cell_sep$lx_groups <- factor(cell_sep$lx_groups, 
                           levels = c("SNP", "scramble", "motif", 
                                      "pos_control"), ordered = T)

```


```{r}
seq_notch = 0.05

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "hmec"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_hmec <-  ggplot(dplyr::filter(cell_sep, celltype == "hmec"), 
                     aes(x = hmec_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("HMEC")
a_hmec

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "airway"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_airway <-  ggplot(dplyr::filter(cell_sep, celltype == "airway"), 
                     aes(x = airway_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
   ylim(start, end) +
  ggtitle("Airway")
a_airway



sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "colon"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]
a_colon <-  ggplot(dplyr::filter(cell_sep, celltype == "colon"), 
                     aes(x = colon_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
   ylim(start, end) +
  ggtitle("Colon")
a_colon

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "pros"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]
a_pros <-  ggplot(dplyr::filter(cell_sep, celltype == "pros"), 
                     aes(x = pros_groups, y = log2(statistic))) +
   geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
   ylim(start, end) +
  ggtitle("Prostate")
a_pros

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "endo"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]
a_endo <-  ggplot(dplyr::filter(cell_sep, celltype == "endo"), 
                     aes(x = endo_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
     ylim(start, end) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ggtitle("Endometrial")
a_endo


grid_p <- grid.arrange(a_hmec, a_airway, a_colon, a_pros, a_endo, 
             left = "log2(alpha)",
             nrow = 1)
ggsave(here("images/figure_panels/mpra_seq_activity_boxplot_yaxis_changed_fig1.pdf"),
       grid_p,
       width = 150, height = 40, units = "mm")
```
Make the rest of the cell types for supplements:
```{r}
seq_notch = 0.05

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "ast"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_ast <-  ggplot(dplyr::filter(cell_sep, celltype == "ast"), 
                     aes(x = ast_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Astrocyte")
a_ast

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "mc"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_mc <-  ggplot(dplyr::filter(cell_sep, celltype == "mc"), 
                     aes(x = mc_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Melanocyte")
a_mc

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "kc"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_kc <-  ggplot(dplyr::filter(cell_sep, celltype == "kc"), 
                     aes(x = kc_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Keratinocyte")
a_kc


sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "eso"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_eso <-  ggplot(dplyr::filter(cell_sep, celltype == "eso"), 
                     aes(x = eso_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Esophageal")
a_eso

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "renal"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_renal <-  ggplot(dplyr::filter(cell_sep, celltype == "renal"), 
                     aes(x = renal_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Renal")
a_renal

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "ov"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_ov <-  ggplot(dplyr::filter(cell_sep, celltype == "ov"), 
                     aes(x = ov_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Ovarian")
a_ov

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "thy"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_thy <-  ggplot(dplyr::filter(cell_sep, celltype == "thy"), 
                     aes(x = thy_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Thyroid")
a_thy

sig_seq = quantile(log2(cell_sep$statistic[cell_sep$celltype == "panc"]), seq(0, 1, seq_notch), na.rm=T)
start = sig_seq[2]
end = sig_seq[length(sig_seq) - 1]

a_panc <-  ggplot(dplyr::filter(cell_sep, celltype == "panc"), 
                     aes(x = panc_groups, y = log2(statistic))) +
  geom_boxplot(outlier.size = 0.001, notch = F, linewidth = 0.1)+
    xlab("") + ylab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 5),
        plot.title = element_text(size = 7, hjust = 0.5)) +
  ylim(start, end) +
  ggtitle("Pancreatic")
a_panc


grid_supp <- grid.arrange(a_ast, a_mc, a_kc, a_eso, a_renal, a_ov, a_thy,a_panc,
             left = "log2(alpha)",
             nrow = 1)
ggsave(here("images/possible_supplementary_panels/mpra_seq_activity_boxplot_supplementary_celltypes.pdf"),
       grid_supp,
       width = 220, height = 40, units = "mm")
```


Do a bar plot of number sig

```{r}
mpraq_plot_dis_filt_w_ctrls <- dplyr::filter(mpraq_plot, celltype == disease |
                                       seq_type_2 %in% c("pos_control", "scramble")) %>%
  rowwise() %>%
  mutate(`Cell type` = case_when(
    seq_type_2 == "pos_control" ~ "Pos Control",
    seq_type_2 == "scramble" ~ "Scramble",
    T ~  name_key$display_name[name_key$celltype == celltype]))

active_bar_plot <- ggplot(mpraq_plot_dis_filt_w_ctrls, 
                          aes(x = `Cell type`, fill = active_seq)) +
  geom_bar() +
  ylab("Number of SNPs") + xlab("") +
  labs(fill = "Activity") +
  scale_fill_manual(values = c("grey80", "black")) +
  theme_minimal() + 
  theme(text = element_text(size = 6, family = "Helvetica"),
    axis.text.x = element_text(angle = 45, vjust = 0.8),
            line=element_line(linewidth=0.2),
        legend.text = element_text(size =5),
        legend.title = element_text(size =6),
        legend.key.size = unit(0.2, 'cm'))



mpraq_plot_dis_filt <- dplyr::filter(mpraq_plot, celltype == disease) %>%
  rowwise() %>%
  mutate(`Cell type` = name_key$display_name[name_key$celltype == celltype])
active_bar_plot <- ggplot(mpraq_plot_dis_filt, 
                          aes(x = `Cell type`, fill = active_seq)) +
  geom_bar() +
  ylab("Number of SNPs") + xlab("") +
  labs(fill = "Activity") +
  scale_fill_manual(values = c("grey80", "black")) +
  theme_minimal() + 
  theme(text = element_text(size = 6, family = "Helvetica"),
    axis.text.x = element_text(angle = 45, vjust = 0.8),
            line=element_line(linewidth=0.2),
        legend.text = element_text(size =5),
        legend.title = element_text(size =6),
        legend.key.size = unit(0.2, 'cm'))



ggsave(here("images/figure_panels/mpra_active_seq_barplots_disease_filt.pdf"),
       active_bar_plot,
      width = 65, height = 40, units = "mm")


```

