---
title: "Untitled"
output: html_document
date: "2023-06-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(ggpubr)
library(rstatix)

```

Make box plots or dot plots for the luciferase and MPRA results to compare
Read in the MPRA counts, normalize for seq depth and to plasmid DNA (just divide)
Read in the luciferase results

Start with HMECs

What do I want?
For each cell type, set of side by side MPRA and luciferase results
For rs55 I want hmec MPRA, airway MPRA, MCF7 luciferase

Go back and fix the MPRA p values so they're MPRAnalyze instead of a t test

Also includes rs10411210 MPRA and luciferase plots

```{r}
res_merge <- read_tsv(here("output/res_merge_withhitannot_20230627.tsv"))
```

```{r}
# supp plot params
fig_height = 30
fig_width = 30
o_size = 0.2
o_stroke = 0.3
l_width = 0.15
v_just = 0
tip_length = 0
p_size = 2
```

HMEC luciferase data

```{r}
brca_luc <- read_tsv(here("data/brca_luciferase_data.txt"))

brca_luc <- brca_luc %>% mutate(snp_allele = paste(SNP, allele, sep="_")) %>%
  rowwise() %>%
  mutate(luc = mean(c(rep1, rep2, rep3, rep4, rep5)))
brca_luc_long <- brca_luc %>% pivot_longer(cols = c(rep1, rep2, rep3, rep4, rep5), names_to = "rep",
                                             values_to = "luc_vals")
brca_luc_long <- brca_luc_long %>%
  mutate(Allele = ifelse(allele == "ref", "Ref", "Alt"))
comparisons <- list(c("Alt", "Ref"))


for (snp in brca_luc$SNP){
  df <- dplyr::filter(brca_luc_long, SNP == snp)
  df_max <- max(df$luc_vals[!is.nan(df$luc_vals)])
  p <- ggplot(df, 
              aes(x = as.factor(Allele), y=luc_vals)) + 
  geom_boxplot(outlier.shape = NA, linewidth = l_width,
                 outlier.size = o_size, outlier.stroke = o_stroke) +
  geom_jitter(width = 0.2, size = o_size, stroke = o_stroke) + theme_minimal() +
  theme(text = element_text(size = 5, family = "Helvetica"),
        plot.title = element_text(size = 5, hjust = 0.5),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 5),
        line=element_line(linewidth=0.2)) +
  ggtitle(paste0(snp, " luciferase activity")) + 
    labs(x = "Allele", y = "Luciferase activity") + 
    stat_compare_means(method = "t.test", comparison = comparisons, label = "p.signif",
                       tip.length = 0.01, vjust = v_just, size = p_size,
                       label.y = df_max + 0.1*df_max) +
        ylim(c(0, df_max + 0.3*df_max))
  
  
  ggsave(here(paste0("images/possible_supplementary_panels/luciferase_plots/brca/",
                     snp, "_luc.pdf")),
         p,
         width = fig_width, height = fig_height, units = "mm")
}

```
Get normalized counts
```{r}
counts <- read_tsv(here("data/nextseq_counts_111920.tsv"))
counts_norm <- counts %>%
  mutate(across(where(is.numeric), ~ ./sum(.))) %>%
  mutate(plasmid = plasmid_2 + plasmid_3) %>%
  mutate(across(where(is.numeric), ~  ./plasmid))

counts_norm <- counts_norm %>%
  separate(oligo, into = c("junk", "locus_num", "allele", "barcode")) %>%
  unite("locus", "junk", "locus_num", sep = "_")

counts_norm <- left_join(dplyr::select(res_merge, locus, Causal_SNP),
          counts_norm) %>%
  mutate(Allele = ifelse(allele == "ref", "Ref", "Alt"))
```

```{r}
brca_counts <- counts_norm %>%
  dplyr::select(locus, Causal_SNP, Allele, barcode, hmec_1, hmec_2, hmec_3) %>%
  pivot_longer(cols = c("hmec_1", "hmec_2", "hmec_3"), 
               names_to = "replicate", values_to = "Activity")

for (snp in brca_luc$SNP){
  df <- dplyr::filter(brca_counts, Causal_SNP == snp)
  df_max <- max(df$Activity[!is.nan(df$Activity)])
  mpra_pval <- data.frame(
    group1 = "Alt",
    group2 = "Ref",
    p = res_merge$hmec_fdr[res_merge$Causal_SNP == snp],
  y.position = df_max + 0.1*df_max,
    xmin = 1,
    xmax = 2)
  mpra_pval <- mpra_pval %>%
    add_significance()
  p <- ggplot(df, 
              aes(x = as.factor(Allele), y=Activity)) + 
    geom_boxplot(outlier.size = o_size, outlier.stroke = o_stroke,
                 linewidth = l_width) +
    theme_minimal() +
    theme(text = element_text(size = 5),
          plot.title = element_text(size = 5),
          axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 5)) +
    ggtitle(paste0(snp, " MPRA activity")) + 
      labs(x = "Allele", y = "MPRA activity") + 
       stat_pvalue_manual(mpra_pval, label = "p.signif", tip.length = 0,
                        vjust = v_just, size = p_size) +
    ylim(c(0, df_max + 0.3*df_max))
  ggsave(here(paste0("images/possible_supplementary_panels/mpra_single_snp_plots/brca/",
                     snp, "_mpra.pdf")),
         p,
         width = fig_width, height = fig_height, units = "mm")

}
```

Make the rs5 panels
```{r eval=FALSE, include=FALSE}
figwidth = 20
figheight = 40
snp = "rs58821592"
# make the MPRA plot

stat.test.hmpra <- data.frame(
  group1 = "Alt",
  group2 = "Ref",
  p = res_merge$hmec_fdr[res_merge$Causal_SNP == snp],
  y.position = 1.25,
  xmin = 1,
  xmax = 2)

stat.test.hmpra <- stat.test.hmpra %>%
  add_significance()

h <- ggplot(dplyr::filter(brca_counts, Causal_SNP == snp), 
              aes(x = as.factor(Allele), y=Activity)) + 
  geom_boxplot(outlier.size = 0.1, linewidth = 0.3) +
    theme_minimal() +
  theme(text = element_text(size = 5, family = "Helvetica"),
        plot.title = element_text(size = 6, hjust = 0.5),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 6),
        line=element_line(linewidth=0.2)) +
  ylim(0, 1.3) +
  ggtitle(paste0(snp, "\nHMEC MPRA")) + 
    labs(x = "Allele", y = "MPRA activity") +
  stat_pvalue_manual(stat.test.hmpra, label = "p.signif", tip.length = 0,
                        vjust = 0.6)

ggsave(here("images/figure_panels/rs58821592_hmec_mpra_plot.pdf"),
         h,
         width = figwidth, height = figheight, units = "mm")


airway_counts <- counts_norm %>%
  dplyr::select(locus, Causal_SNP, Allele, barcode, airway_1, airway_2, airway_3) %>%
  pivot_longer(cols = c("airway_1", "airway_2", "airway_3"), 
               names_to = "replicate", values_to = "Activity")
a <- ggplot(dplyr::filter(airway_counts, Causal_SNP == snp), 
              aes(x = as.factor(Allele), y=Activity)) + 
  geom_boxplot(outlier.size = 0.1, linewidth = 0.3) +
    theme_minimal() +
  theme(text = element_text(size = 5, family = "Helvetica"),
        plot.title = element_text(size = 6, hjust = 0.5),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 6),
        line=element_line(linewidth=0.2)) +
  ylim(0, 2) +
  ggtitle(paste0(snp, "\nAirway MPRA")) + 
    labs(x = "Allele", y = "MPRA activity")
ggsave(here("images/figure_panels/rs58821592_airway_mpra_plot.pdf"),
         a,
         width = figwidth, height = figheight, units = "mm")

hl <- ggplot(dplyr::filter(brca_luc_long, SNP == snp), 
              aes(x = as.factor(Allele), y=luc_vals)) + 
  geom_boxplot(outlier.shape = NA, linewidth = 0.3) +
  geom_jitter(width = 0.2, size = 0.1) + theme_minimal() +
  theme(text = element_text(size = 5, family = "Helvetica"),
        plot.title = element_text(size = 6, hjust = 0.8),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 6),
        line=element_line(linewidth=0.2)) +
  ggtitle(paste0(snp, "\nluciferase in MCF7")) + 
    labs(x = "Allele", y = "Luciferase activity") + 
  ylim(0, 1.5) +
    stat_compare_means(method = "t.test", comparison = comparisons, label = "p.signif",
                       tip.length = 0.01, vjust = 0, size = 2)
  
  ggsave(here("images/figure_panels/rs58821592_luciferase_plot.pdf"),
         hl,
         width = figwidth, height = figheight, units = "mm")

```

Colon luciferase data
```{r}
colon_luc <- read_tsv(here("data/colon_luciferase_data.txt"))
colon_luc_long <- colon_luc %>% pivot_longer(cols = c(rep1, rep2, rep3, rep4, rep5), names_to = "rep",
                                             values_to = "luc_vals")
colon_luc_long <- colon_luc_long %>%
  mutate(Allele = ifelse(allele == "ref", "Ref", "Alt"))
comparisons <- list(c("Alt", "Ref"))

for (snp in colon_luc$SNP){
  df <- dplyr::filter(colon_luc_long, SNP == snp)
  df_max <- max(df$luc_vals[!is.nan(df$luc_vals)])
  p <- ggplot(df, 
              aes(x = as.factor(Allele), y=luc_vals)) + 
    geom_boxplot(outlier.shape = NA, linewidth = l_width,
                 outlier.size = o_size, outlier.stroke = o_stroke) +
  geom_jitter(width = 0.2, size = o_size, stroke = o_stroke) + theme_minimal() +
  theme(text = element_text(size = 5, family = "Helvetica"),
        plot.title = element_text(size = 5, hjust = 0.5),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 5),
        line=element_line(linewidth=0.2)) +
  ggtitle(paste0(snp, " luciferase activity")) + 
    labs(x = "Allele", y = "Luciferase activity") + 
    stat_compare_means(method = "t.test", comparison = comparisons, label = "p.signif",
                       tip.length = 0.01, vjust = v_just, size = p_size,
                       label.y = df_max + 0.1*df_max) +
        ylim(c(0, df_max + 0.3*df_max))
  
  ggsave(here(paste0("images/possible_supplementary_panels/luciferase_plots/colon/",
                     snp, "_luc.pdf")),
         p,
         width = fig_width, height = fig_height, units = "mm")
}

```

Colon MPRA plots
```{r}
colon_counts <- counts_norm %>%
  dplyr::select(locus, Causal_SNP, Allele, barcode, colon_1, colon_2, colon_3) %>%
  pivot_longer(cols = c("colon_1", "colon_2", "colon_3"), 
               names_to = "replicate", values_to = "Activity")

for (snp in colon_luc$SNP){
  df <- dplyr::filter(colon_counts, Causal_SNP == snp)
  df_max <- max(df$Activity[!is.nan(df$Activity)])
  mpra_pval <- data.frame(
    group1 = "Alt",
    group2 = "Ref",
    p = res_merge$colon_fdr[res_merge$Causal_SNP == snp],
    y.position = df_max + 0.1*df_max,
    xmin = 1,
    xmax = 2)
  mpra_pval <- mpra_pval %>%
    add_significance()
  p <- ggplot(dplyr::filter(colon_counts, Causal_SNP == snp), 
              aes(x = as.factor(Allele), y=Activity)) + 
    geom_boxplot(linewidth = l_width,  outlier.size = o_size, 
                 outlier.stroke = o_stroke) +
    theme_minimal() +
  theme(text = element_text(size = 5),
        plot.title = element_text(size = 5),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 5)) +
  ggtitle(paste0(snp, " MPRA activity")) + 
    labs(x = "Allele", y = "MPRA activity") + 
    stat_pvalue_manual(mpra_pval, label = "p.signif", tip.length = 0,
                        vjust = v_just, size = p_size) +
        ylim(c(0, df_max + 0.3*df_max))

  ggsave(here(paste0("images/possible_supplementary_panels/mpra_single_snp_plots/colon/",
                     snp, "_mpra.pdf")),
         p,
         width = fig_width, height = fig_height, units = "mm")
}
```

Make the rs10411210 panels
```{r eval=FALSE, include=FALSE}
rs1_figwidth = 30
rs1_figheight = 45
snp = "rs10411210"

rs1_df <- dplyr::filter(colon_counts, Causal_SNP == snp)
rs1_df <- rs1_df %>% mutate(Allele = ifelse(Allele == "Alt", "T (Protective)",
                                 "C (Risk)"))
rs1_df$Allele <- factor(rs1_df$Allele, levels = c("C (Risk)", "T (Protective)"),
                        ordered = T)


stat.test.rs1colon <- data.frame(
  group1 = "Alt",
  group2 = "Ref",
  p = res_merge$colon_fdr[res_merge$Causal_SNP == snp],
  y.position = 1.7,
  xmin = 1,
  xmax = 2)

stat.test.rs1colon <- stat.test.rs1colon %>%
  add_significance()


# make the MPRA plot
r1_m <- ggplot(rs1_df, 
              aes(x = as.factor(Allele), y=Activity)) + 
  geom_boxplot(outlier.size = o_size, 
                 outlier.stroke = o_stroke,
               linewidth = l_width) +
    theme_minimal() +
  theme(text = element_text(size = 5, family = "Helvetica"),
        plot.title = element_text(size = 6, hjust = 0.5),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 5),
        line=element_line(linewidth=0.2)) +
  ggtitle(paste0(snp, "\nColon MPRA")) + 
    labs(x = "Allele", y = "MPRA activity") +
    stat_pvalue_manual(stat.test.rs1colon, label = "p.signif", tip.length = 0,
                        vjust = 0.4, size = 2) +
  ylim(c(0, max(rs1_df$Activity)*1.1))
r1_m
ggsave(here("images/figure_panels/rs10411210_colon_mpra_plot.pdf"),
         r1_m,
         width = rs1_figwidth, height = rs1_figheight, units = "mm")

rs1_luc_df <- dplyr::filter(colon_luc_long, SNP == snp)
rs1_luc_df <- rs1_luc_df %>% mutate(Allele = ifelse(allele == "alt", "T (Protective)",
                                 "C (Risk)"))
rs1_luc_df$Allele <- factor(rs1_luc_df$Allele, levels = c("C (Risk)", 
                                                          "T (Protective)"),
                        ordered = T)
allele_comp = list(c("C (Risk)", "T (Protective)"))

r1_l <- ggplot(rs1_luc_df, 
              aes(x = Allele, y=luc_vals)) + 
  geom_boxplot(outlier.size = o_size, 
                 outlier.stroke = o_stroke,
               linewidth = l_width) +
    theme_minimal() +
  theme(text = element_text(size = 5, family = "Helvetica"),
        plot.title = element_text(size = 6, hjust = 0.5),
        axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45, size = 5),
        line=element_line(linewidth=0.2)) +
  ggtitle(paste0(snp, "\nluciferase in HCT116")) + 
    labs(x = "Allele", y = "Luciferase activity") + 
    stat_compare_means(method = "t.test", comparison = allele_comp, label = "p.signif",
                       tip.length = 0.0, vjust = .4, size = 2) +
  ylim(c(0, max(rs1_luc_df$luc_vals) * 1.1))
r1_l
  
  ggsave(here("images/figure_panels/rs10411210_HCT116_luciferase_plot.pdf"),
         r1_l,
         width = rs1_figwidth, height = rs1_figheight, units = "mm")

```
